The Courtroom Challenge—A Conjecture on the Unavoidable Cost of LLM Verification

### 2025-07-28 main.tex
Add files via upload
Sparse-memory tightness + bridge to conjectures

* Lemma 3 (“Stochastic tightness of τ-packing”)

* New discussion paragraph after Lemma 3

% =========================================================
% main.tex  –  arXiv-ready article template (CS categories)
% =========================================================
\documentclass[11pt]{article}

@@

% ---------- bibliography ----------
% Author–year (ACL-like)
\usepackage[
  backend=biber,
  style=authoryear,
  natbib=true,      % enables \citep / \citet
  maxbibnames=6,
  minbibnames=1,
  maxcitenames=2,
  sorting=none
]{biblatex}
\addbibresource{refs.bib}

@@

% ---------- quick-reference box ------------------------------------
\begin{tcolorbox}[colback=black!3,colframe=black!40,
  boxrule=0.4pt, arc=2pt, left=2pt, right=2pt,
  top=4pt, bottom=4pt, title=Notation snapshot]
\small
\begin{itemize}[leftmargin=*]
\item $G$ – \textbf{Generator}: LLM (+ optional helper calls)
that proposes a hypothesis $h_t$ for query $q_t$.
\item $V_{\theta}$ – \textbf{Validator} with
\emph{validation–preference dial} $\theta\!\in[0,1]$  
($0 =\,$speed‐first, $1 =\,$accuracy‐first).  
The dial is fixed during any evaluation run; see
Sec.~\ref{sec:framework}.
\item $\kappa_i c_{i,t}$ – \textbf{Cost-ledger term}: each resource $i$
(bytes written, CPU‐seconds, human minutes …) is logged as a
quantity $c_{i,t}$ weighted by its unit cost $\kappa_i$.
$\sum_i\kappa_i c_{i,t}$ is the friction paid at step $t$.
\item $\tau$ – \textbf{Retention threshold} that controls whether a
new descriptor vector is written to the precedent store:  
\vspace{-2pt}
\[
\text{write\_new}=\begin{cases}
[\cos\theta < \tau], &\text{(cosine form)}\\[4pt]
[\lVert x-y\rVert_2 > \tau], &\text{(distance form)}
\end{cases}
\]
Designers may substitute any monotone similarity metric.
\end{itemize}
\vspace{2pt}\par\footnotesize (Formal definitions → Section~\ref{sec:framework})\par
\end{tcolorbox}
% --------------------------------------------------------------------

@@

Cosine and L$_2$ are special cases, but a system designer is free to
plug in any monotone metric—e.g.\ Mahalanobis distance, dot-product in a
learned embedding space, or even a hybrid score that blends semantic and
numeric keys—so long as the inequality is flipped to match the
metric's notion of proximity. Section~\ref{sec:sparse-memory-bound} analyses the cost/error trade-off
under the distance form (Lemma~\ref{lemma}), and Appendix~\ref{sec:knobs} lists this dial
among the falsifiability-critical degrees of freedom. Selecting which descriptor 
fields to embed is itself tunable; Section~\ref{sec:open-questions} (Q\ref{open-question-schema}) sketches how the Courtroom loop 
could learn or prune schema columns over time.

@@

\section{Architectural Analogues}
\label{sec:analogues}
Structured exploration, exemplified by AlphaGo's policy–value search \citep{silver2016alphago} and more recently by Tree-of-Thoughts planning \citep{yao2023tree}, offers a compelling model for integrating creativity with systematic validation. In AlphaGo, a policy network generates moves, a value network evaluates outcomes, and self-play iteratively refines strategy. The Courtroom Model generalizes this principle to open-ended domains, applying continuous validation not only internally but also against a persistent external memory and, when needed, human oversight. This convergence supports the idea that the Courtroom Model is not prescriptive, but descriptive — a lens for recognizing effective reasoning architectures wherever they emerge. It offers a retrospective framework to describe such structures — and perhaps guide the design of future systems that balance creativity with constraint.

@@

  \item \textbf{Validator family.}  
  \(V_{\theta} : (Q, H, M) \to \{\textsc{ACCEPT}, \textsc{REJECT}, \textsc{ESCALATE}, \textsc{DEFER}\}\), with a
  tunable validation–preference dial \(\theta \in \Theta\).
  \(V_{\theta}\) may be
  (i) a stronger ensemble LLM,
  (ii) a formal-proof kernel,
  (iii) a deterministic numeric or symbolic tool-chain that certifies or refutes the claim, or
  (iv) a human panel as fallback.\footnote{\textit{Self-play precedent.}  
  AlphaZero's value network functions as an automated judge for Monte-Carlo rollouts.
  The Courtroom validator is its open-domain analogue: any module that can reliably
  downgrade faulty claims can occupy the judge slot.} \textsc{DEFER} means “skip this round”: no answer is returned,
  no escalation cost is paid, and the generator proceeds to the next
  query.

@@

\paragraph{Provisionality of verdicts.}
Even the external Judge’s verdicts are provisional: they may be rescinded
if subsequent audits or new primary filings contradict earlier
conclusions.  Accordingly, Judge-validated precedents can be re-queued,
incur additional \(\WriteCost\), and overwrite prior entries via the usual
supersession mechanism.

@@

Given $S_t=(q_t,h_t,a_t,M_t)$:
\begin{enumerate}
  \item $h_t \leftarrow G(q_t,M_t)$
  \item $d_t \leftarrow V_\theta(q_t,h_t,M_t)$
        \begin{itemize}
          \item If $d_t=\mathrm{ACCEPT}$ set $a_t=h_t$
          \item Else obtain corrected answer $a_t = C(q_t,h_t)$
                (oracle $C$ may call a human; cost is logged)
          \item If $d_t=\textsc{DEFER}$ set $a_t=\bot$
                and skip Steps 3–4 below.
        \end{itemize}
  \item $M_{t+1} \leftarrow U(M_t,q_t,h_t,a_t)$
  \item External stream provides $q_{t+1}$
\end{enumerate}

@@

\paragraph{Per-round notation.}
Fix a margin parameter $\delta\in(0,1]$.
For every run $\omega$ and round $t=1,\dots,T$ set
\begin{align*}
  y_t      &:=y_t(\theta)          &&\text{(ground-truth label)},\\
  h_t      &:=h_t(\omega)          &&\text{(validator’s automatic prediction)},\\
  \mathrm{Esc}_t&:=\mathbf 1_{\{\text{round $t$ is escalated}\}.}
\end{align*}
Define the \emph{leverage rounds}
\begin{equation}\label{eq:kappaDef}
  \mathcal L(\omega):=\Bigl\{\,t:\,
     \mathrm{Esc}_t=0\;\text{and}\;|y_t-h_t|\ge\delta\Bigr\},
  \qquad
  \kappa(\omega):=\frac{|\mathcal L(\omega)|}{T}.
\end{equation}
The cumulative label error on those rounds is
\begin{equation}\label{eq:ErrDef}
  \Err(\omega):=\sum_{t\in\mathcal L(\omega)}
                \mathbf 1_{\!\bigl[h_t\neq y_t\bigr]}.
\end{equation}
Define the global means:
\[
  \bar\kappa:=\E[\kappa],
  \qquad
  \Err_{\mathrm{bar}}:=\E[\Err].
\]

@@

\subsection{Sparse-memory bound}
\label{sec:sparse-memory-bound}

The following lemma gives a back-of-the-envelope bound on memory growth and justifies the choice of a retention threshold $\tau$.

\begin{lemma}
\label{lemma}
Assume (i) queries are i.i.d.\ in a bounded metric space
and (ii)  $U$ inserts a new vector only if its L$_2$ distance to every
vector in $M_t$ exceeds a threshold~$\tau$.  Then
\[
  |M_T| \;\le\;
  \bigl(\tfrac{\operatorname{diam}(Q)}{\tau}\bigr)^{d},
  \qquad
  \forall\,T\ge 0 .
\]
Hence total storage cost is $O\!\bigl(\tau^{-d}\bigr)$.
\end{lemma}

\begin{proof}[Proof sketch]
The insertion rule builds an $\tau$-packing of $Q$; a standard $\varepsilon$-net
argument bounds the packing number by $(\mathrm{diam}/\tau)^d$.
\end{proof}

>>>>

% =========================================================
% main.tex  -  arXiv-ready article template (CS categories)
% =========================================================
\documentclass[11pt]{article}

@@

% ---------- bibliography ----------
% Author-year (ACL-like)
\usepackage[
  backend=biber,
  style=authoryear,
  natbib=true,      % enables \citep / \citet
  maxbibnames=6,
  minbibnames=1,
  maxcitenames=2,
  sorting=none
]{biblatex}
\addbibresource{refs.bib}

@@

% ---------- quick-reference box ------------------------------------
\begin{tcolorbox}[colback=black!3,colframe=black!40,
  boxrule=0.4pt, arc=2pt, left=2pt, right=2pt,
  top=4pt, bottom=4pt, title=Notation snapshot]
\small
\begin{itemize}[leftmargin=*]
\item $G$ - \textbf{Generator}: LLM (+ optional helper calls)
that proposes a hypothesis $h_t$ for query $q_t$.
\item $V_{\theta}$ - \textbf{Validator} with
\emph{validation-preference dial} $\theta\!\in[0,1]$  
($0 =\,$speed‐first, $1 =\,$accuracy‐first).  
The dial is fixed during any evaluation run; see
Sec.~\ref{sec:framework}.
\item $\kappa_i c_{i,t}$ - \textbf{Cost-ledger term}: each resource $i$
(bytes written, CPU‐seconds, human minutes …) is logged as a
quantity $c_{i,t}$ weighted by its unit cost $\kappa_i$.
$\sum_i\kappa_i c_{i,t}$ is the friction paid at step $t$.
\item $\tau$ - \textbf{Retention threshold} that controls whether a
new descriptor vector is written to the precedent store:  
\vspace{-2pt}
\[
\text{write\_new}=\begin{cases}
[\cos\theta < \tau], &\text{(cosine form)}\\[4pt]
[\lVert x-y\rVert_2 > \tau], &\text{(distance form)}
\end{cases}
\]
Designers may substitute any monotone similarity metric.
\end{itemize}
\vspace{2pt}\par\footnotesize (Formal definitions → Section~\ref{sec:framework})\par
\end{tcolorbox}
% --------------------------------------------------------------------

@@

Cosine and L$_2$ are special cases, but a system designer is free to
plug in any monotone metric—e.g.\ Mahalanobis distance, dot-product in a
learned embedding space, or even a hybrid score that blends semantic and
numeric keys—so long as the inequality is flipped to match the
metric's notion of proximity. Section~\ref{sec:sparse-memory-bound-tightness} analyses the cost/error trade-off
under the distance form (Lemma~\ref{lemma}), and Appendix~\ref{sec:knobs} lists this dial
among the falsifiability-critical degrees of freedom. Selecting which descriptor 
fields to embed is itself tunable; Section~\ref{sec:open-questions} (Q\ref{open-question-schema}) sketches how the Courtroom loop 
could learn or prune schema columns over time.

@@

\section{Architectural Analogues}
\label{sec:analogues}
Structured exploration, exemplified by AlphaGo's policy-value search \citep{silver2016alphago} and more recently by Tree-of-Thoughts planning \citep{yao2023tree}, offers a compelling model for integrating creativity with systematic validation. In AlphaGo, a policy network generates moves, a value network evaluates outcomes, and self-play iteratively refines strategy. The Courtroom Model generalizes this principle to open-ended domains, applying continuous validation not only internally but also against a persistent external memory and, when needed, human oversight. This convergence supports the idea that the Courtroom Model is not prescriptive, but descriptive — a lens for recognizing effective reasoning architectures wherever they emerge. It offers a retrospective framework to describe such structures — and perhaps guide the design of future systems that balance creativity with constraint.

@@

  \item \textbf{Validator family.}  
  \(V_{\theta} : (Q, H, M) \to \{\textsc{ACCEPT}, \textsc{REJECT}, \textsc{ESCALATE}, \textsc{DEFER}\}\), with a
  tunable validation-preference dial \(\theta \in \Theta\).
  \(V_{\theta}\) may be
  (i) a stronger ensemble LLM,
  (ii) a formal-proof kernel,
  (iii) a deterministic numeric or symbolic tool-chain that certifies or refutes the claim, or
  (iv) a human panel as fallback.\footnote{\textit{Self-play precedent.}  
  AlphaZero's value network functions as an automated judge for Monte-Carlo rollouts.
  The Courtroom validator is its open-domain analogue: any module that can reliably
  downgrade faulty claims can occupy the judge slot.} \textsc{DEFER} means “skip this round”: no answer is returned,
  no escalation cost is paid, and the generator proceeds to the next
  query.

@@

\paragraph{Provisionality of verdicts.}
Even the external Judge's verdicts are provisional: they may be rescinded
if subsequent audits or new primary filings contradict earlier
conclusions.  Accordingly, Judge-validated precedents can be re-queued,
incur additional \(\WriteCost\), and overwrite prior entries via the usual
supersession mechanism.

@@

Given $S_t=(q_t,h_t,a_t,M_t)$:
\begin{enumerate}
  \item $h_t \leftarrow G(q_t,M_t)$
  \item $d_t \leftarrow V_\theta(q_t,h_t,M_t)$
        \begin{itemize}
          \item If $d_t=\mathrm{ACCEPT}$ set $a_t=h_t$
          \item Else obtain corrected answer $a_t = C(q_t,h_t)$
                (oracle $C$ may call a human; cost is logged)
          \item If $d_t=\textsc{DEFER}$ set $a_t=\bot$
                and skip Steps 3-4 below.
        \end{itemize}
  \item $M_{t+1} \leftarrow U(M_t,q_t,h_t,a_t)$
  \item External stream provides $q_{t+1}$
\end{enumerate}

@@

\paragraph{Per-round notation.}
Fix a margin parameter $\delta\in(0,1]$.
For every run $\omega$ and round $t=1,\dots,T$ set
\begin{align*}
  y_t      &:=y_t(\theta)          &&\text{(ground-truth label)},\\
  h_t      &:=h_t(\omega)          &&\text{(validator's automatic prediction)},\\
  \mathrm{Esc}_t&:=\mathbf 1_{\{\text{round $t$ is escalated}\}.}
\end{align*}
Define the \emph{leverage rounds}
\begin{equation}\label{eq:kappaDef}
  \mathcal L(\omega):=\Bigl\{\,t:\,
     \mathrm{Esc}_t=0\;\text{and}\;|y_t-h_t|\ge\delta\Bigr\},
  \qquad
  \kappa(\omega):=\frac{|\mathcal L(\omega)|}{T}.
\end{equation}
The cumulative label error on those rounds is
\begin{equation}\label{eq:ErrDef}
  \Err(\omega):=\sum_{t\in\mathcal L(\omega)}
                \mathbf 1_{\!\bigl[h_t\neq y_t\bigr]}.
\end{equation}
Define the global means:
\[
  \bar\kappa:=\E[\kappa],
  \qquad
  \Err_{\mathrm{bar}}:=\E[\Err].
\]

@@

\subsection{Sparse-memory bounds and tightness}
\label{sec:sparse-memory-bound-tightness}

We bound how many precedents a validator stores when it keeps only queries that
are at least a distance~$\tau$ apart.  The first lemma gives a worst-case upper
bound; the second shows that bound is tight under mild stochastic traffic.

\begin{lemma}[Sparse-memory upper bound]\label{lem:2}
Let $\mathcal Q\subset\mathbb R^{d}$ be bounded with
$\operatorname{diam}\mathcal Q=D<\infty$.
Suppose the validator maintains a set $M_t$ that starts empty and, at each
round $t$, inserts the query $q_t$ \emph{only if}
\(\min_{m\in M_{t-1}}\lVert q_t-m\rVert_2>\tau\).
Then for all $T\ge0$
\[
  |M_T|
  \;\le\;
  \Bigl(\tfrac{\operatorname{diam}\mathcal Q}{\tau}\Bigr)^{d}.
\]
Hence total storage cost is \(O\!\bigl(\tau^{-d}\bigr)\).
\end{lemma}

\begin{proof}[Proof sketch]
The rule ensures $M_T$ is a $\tau$-packing of $\mathcal Q$.
By a standard $\varepsilon$-net/volume argument, any $\tau$-packing inside a
$D$-diameter set has size at most $(D/\tau)^{d}$.
\end{proof}

\begin{lemma}[Stochastic tightness of $\tau$-packing]\label{lem:3}
Let the setting of Lemma \ref{lem:2} hold, and assume in addition that the
queries $(q_t)_{t\ge1}$ are i.i.d.\ with a density
$f\ge\lambda>0$ on $\mathcal Q$.
Define
\[
  T_0(\tau,\delta)\;=\;
  \frac{2^{d}}{\lambda\,\nu_d}\,
  \tau^{-d}\,
  \Bigl[d\log\!\bigl(D/\tau\bigr)+\log(1/\delta)\Bigr],
  \qquad
  \nu_d=\frac{\pi^{d/2}}{\Gamma(d/2+1)}.
\]
Then, for any confidence level $0<\delta<1$ and all $T\ge T_0$,
\[
  \Pr\!\Bigl[\,|M_T|\;\ge\;
        \tfrac12\,\nu_d^{-1}D^{-d}\,\tau^{-d}\Bigr]
  \;\ge\;1-\delta.
\]
In particular, under full-support i.i.d.\ traffic the memory size scales
\(\Theta(\tau^{-d})\) with high probability.
\end{lemma}

\begin{proof}[Sketch]
1.~Maximal $\tau$-packing.
Let $\{c_i\}_{i=1}^{m}\subset\mathcal Q$ be maximal with pairwise distances
$>\tau$.  By inscribing disjoint axis-aligned cubes of side $\tau/\sqrt d$ we
obtain $m\ge\alpha_d(D/\tau)^d$ with $\alpha_d=2^{-d}$.

\noindent
2.~Disjoint balls.
Set $B_i=B(c_i,\tfrac{\tau}{2})$.  The balls are disjoint and lie inside
$\mathcal Q$.

\noindent
3.~Uniform hit probability.
By the density lower bound, each ball has
$p_\tau=\lambda\nu_d2^{-d}\tau^{d}$
probability mass; the value is the same for every $i$.

\noindent
4.~Empty-ball tail.
$\Pr(B_i\text{ empty})=(1-p_\tau)^T\le e^{-p_\tau T}$
(using $\log(1-x)\le -x$).  Choosing $T\ge T_0$
forces $e^{-p_\tau T}\le\delta/m$.

\noindent
5.~Union bound and insertion rule.
With probability $\ge1-\delta$ every $B_i$ is hit at least once.
The first hit in each ball is $>\tau$ from all stored points,
hence inserted; so $|M_T|\ge m\ge\alpha_d(D/\tau)^d$.
Taking $c_d=\tfrac12\nu_d^{-1}D^{-d}\le\alpha_d$ matches the lemma's constant.
\end{proof}

\paragraph{Discussion — from tightness to the trilemma.}
Lemma~\ref{lem:2} caps memory by $O(\tau^{-d})$ for any query
sequence.  Lemma~\ref{lem:3} shows that, under full-support
i.i.d.\ traffic, the same $\Theta(\tau^{-d})$ rate is unavoidable.
Together with the Fano-style information bound in
Lemma~\ref{lem:ErrLower} we now have two orthogonal “laws of friction”:

\begin{itemize}\setlength\itemsep{2pt}
\item[\textbullet] Information ledger.
  To keep transcripts short, the validator must tolerate a
  Fano error floor.  
\item[\textbullet] Memory ledger.
  At granularity $\tau$ it must provision
  $\Theta(\tau^{-d})$ precedent cells—no compression trick beats this
  when queries explore the space fully.
\end{itemize}

\noindent
What remains open—and is precisely what the next two conjectures tackle—is a
joint lower bound that couples both ledgers plus the escalation
(human-minutes) ledger when queries arrive in an open-world, potentially
adaptive fashion.  The conjectures formulate that missing link.
